---
title: "Bike Buyers Analysis"
author: "Arrhat Maharjan"
date: "2025-03-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Task 1. Data Cleaning and Preparation

In this section, we'll clean the bike buyers dataset and prepare it for analysis. This includes handling missing values, checking for outliers, and ensuring proper data types.

```{r data_preparation}
library(ggplot2)
library(corrplot)

# task 1 - data cleaning and preparation
# load the dataset
bike_buyers.dataset <- read.csv("./bike_buyers.csv", stringsAsFactors = TRUE)

# replace the empty value with NA
bike_buyers.dataset[bike_buyers.dataset == ""] <- NA

# make a copy of the dataset
dataset <- bike_buyers.dataset

# structure of the dataser
str(dataset)

#summary of the dataset
summary(dataset)

#check number of NAs
colSums(is.na(dataset))

# omit the row with any NA values
dataset <- na.omit(dataset)

# drop unused factor levels
dataset <- droplevels(dataset)

#check number of NAs
colSums(is.na(dataset))

# save cleaned dataset
write.csv(dataset, "new_data.csv", row.names = FALSE)

# structure of cleaned dataset
str(dataset)

```

The structure stays the same after cleaning the data.

The data cleaning process involved:

1. Converting empty strings to NA values
2. Removing rows with any NA values
3. Dropping unused factor levels
4. Saving the cleaned dataset for future use

\pagebreak

## Task 2. Summary of Variables

In this section, we'll look at the variables in our dataset using descriptive statistics and visualizations to understand the data.

**Summary of Variables:**

**ID (Numerical)**: Unique identifier (not useful for analysis). 952 entries after cleaning the data.

**Marital.Status (Categorical)**: Represents the marital status of an individual. 518 Married individuals, 434 Single.

```{r marital_status_pie}
# define color pallette for our visual representation
colors <- c("skyblue", "coral1","darkseagreen","mediumpurple","darkorange1")

# pie chart marital status distribution
marital_count <- table(dataset$Marital.Status)
marital_percent <- round(100 * marital_count / sum(marital_count), 1)
martial_label <- paste(names(marital_count), "\n", marital_percent, "%")
pie(marital_count,
    labels = martial_label,
    col = colors,
    main = "Marital Status Distribution")
```

\pagebreak

**Gender (Categorical)**: Represents the gender of the individual. 473 Females, 479 Males.

```{r gender_pie}
# pie chart gender distribution
gender_count <- table(dataset$Gender)
gender_percent <- round(100 * gender_count / sum(gender_count), 1)
gender_label <- paste(names(gender_count), "\n", gender_percent, "%")
pie(gender_count,
    labels = gender_label,
    col = colors,
    main = "Gender Distribution")
```

Marital.Status and Gender have a fairly balanced distribution with marital (49.7% Female, 50.3% Male) and marital status (54.4% Married, 45.6% Single).

\pagebreak

**Income (Numerical)**: Annual income of the individual (continuous). Ranges from $10,000 to $170,000, with a median of $60,000 and mean $55,903.

```{r income_hist}
# histogram income distribution
ggplot(dataset, aes(x = Income)) +
  geom_histogram(
    binwidth = 10000,
    fill = "skyblue",
    color = "black",
    alpha = 1
  ) +
  labs(title = "Income Distribution", x = "Income", y = "Frequency")
```

\pagebreak

**Children (Numerical)**: Number of children (discrete). Minimum 0, maximum 5, with an average of 1.89 children per household.

```{r children_boxplot}
# box plot children per household
ggplot(dataset, aes(x = "", y = Children)) +
  geom_boxplot(fill = "skyblue",
               color = "black",
               width = 0.2) +
  geom_hline(
    aes(yintercept = mean(Children)),
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(title = "Children", y = "Number of Children") +
  theme(axis.title.x = element_blank())
```

\pagebreak

**Education (Categorical)**: Indicates the highest level of education completed. Most individuals have a Bachelor's degree (30.7%/292) or Partial College education (26.5%/252).

```{r education_bar}
# bar chart education distribution
ggplot(dataset, aes(x = Education)) +
  geom_bar(fill = "skyblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Education Distribution", x = "Education", y = "Count")
```

\pagebreak

**Occupation (Categorical)**: Job category of the individual.Professionals (27.6%/263) and Skilled Manual workers (25.4%/242) are the most common occupations.

```{r occupation_bar}
# bar chart occupation distribution
ggplot(dataset, aes(x = Occupation)) +
  geom_bar(fill = "skyblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Occupation Distribution", x = "Occupation", y = "Count")
```

\pagebreak

**Home.Owner (Categorical)**: Indicates whether the individual owns a home. Most individuals (68.5%/652) own their homes while the rest(300) do not.

```{r home_owner_pie}
# pie chart for home ownership
home_owner_count <- table(dataset$Home.Owner)
home_owner_percent <- round(100 * home_owner_count / sum(home_owner_count), 1)
home_owner_label <- paste(names(home_owner_count), "\n", home_owner_percent, "%")
pie(home_owner_count,
    labels = home_owner_label,
    col = colors,
    main = "Home Ownership Distribution")
```

\pagebreak

**Cars (Numerical)**: Number of cars owned (discrete). Most individuals own 1 or 2 cars, with a maximum being 4.

```{r cars_bar}
# bar chart for number of cars owned
ggplot(dataset, aes(x = factor(Cars))) +
  geom_bar(fill = "skyblue") +
  labs(title = "Distribution of Cars Owned", 
       x = "Number of Cars", 
       y = "Count") +
  theme_minimal()
```

\pagebreak

**Commute.Distance (Categorical)**: Represents how far the person commutes daily. Most individuals (35.6%/339) have a short commute of 0-1 miles, followed by 5-10 miles (19.1%/182).

```{r commute_pie}
# pie chart for commute distance
commute_count <- table(dataset$Commute.Distance)
commute_percent <- round(100 * commute_count / sum(commute_count), 1)
commute_label <- paste(names(commute_count), "\n", commute_percent, "%")
pie(commute_count,
    labels = commute_label,
    col = colors,
    main = "Commute Distance Distribution")
```

\pagebreak

**Region (Categorical)**: The geographical region where the individual lives. The dataset is skewed toward North America (51.1%/486), with Europe (30.0%/286) and Pacific (18.9%/180) having fewer representatives.

```{r region_pie}
# pie chart for region
region_count <- table(dataset$Region)
region_percent <- round(100 * region_count / sum(region_count), 1)
region_label <- paste(names(region_count), "\n", region_percent, "%")
pie(region_count,
    labels = region_label,
    col = colors,
    main = "Regional Distribution")
```

\pagebreak

**Age (Numerical)**: Age of the individual (continuous). Ranges from 25 to 89, with a median age of 43 and mean of 44.26. Majority of the individuals in middle-age.

```{r age_hist}
# histogram age distribution
ggplot(dataset, aes(x = Age)) +
  geom_histogram(
    binwidth = 5,
    fill = "skyblue",
    color = "black",
    alpha = 1
  ) +
  labs(title = "Age Distribution", x = "Age", y = "Frequency")
```

\pagebreak

**Purchased.Bike (Categorical)**: Indicates whether the individual purchased a bike.The data shows a fairly balanced distribution with 47.9%/456 of individuals having purchased a bike and 52.1%/496 not having done so.

```{r bike_purchase_pie}
# pie chart for bike purchase
bike_purchase_count <- table(dataset$Purchased.Bike)
bike_purchase_percent <- round(100 * bike_purchase_count / sum(bike_purchase_count), 1)
bike_purchase_label <- paste(names(bike_purchase_count), "\n", bike_purchase_percent, "%")
pie(bike_purchase_count,
    labels = bike_purchase_label,
    col = colors,
    main = "Bike Purchase Distribution")
```

\pagebreak


```{r variable_summary_dataframe, echo=FALSE}

# summary table
variable_summary <- data.frame(
  Variable = c("ID", "Marital Status", "Gender", "Income", "Children", 
               "Education", "Occupation", "Home Owner", "Cars", 
               "Commute Distance", "Region", "Age", "Purchased Bike"),
  Description = c(
    "952 entries after cleaning the data",
    "518 Married (54.4%), 434 Single (45.6%)",
    "473 Females (49.7%), 479 Males (50.3%)",
    "Range: $10,000-$170,000, Median: $60,000, Mean: $55,903",
    "Range: 0-5, Mean: 1.89 children per household",
    "Most common: Bachelors (30.7%), Partial College (26.5%)",
    "Most common: Professional (27.6%), Skilled Manual (25.4%)",
    "68.5% own a home, 31.5% do not",
    "Most own 1 or 2 cars, maximum: 4",
    "Most common: 0-1 Miles (35.6%), 5-10 Miles (19.1%)",
    "North America: 51.1%, Europe: 30.0%, Pacific: 18.9%",
    "Range: 25-89 years, Median: 43, Mean: 44.26",
    "47.9% purchased a bike, 52.1% did not"
  )
)

# Display the summary table
knitr::kable(variable_summary, caption = "Summary of Variables")
```

\pagebreak

## Task 3. Income Analysis

### a. Income Distribution and Statistics

```{r income_summary}
# summary statistics income
summary_stats <- data.frame(
  Mean = mean(dataset$Income),
  Median = median(dataset$Income),
  Variance = var(dataset$Income),
  SD = sd(dataset$Income)
)
summary_stats
```

The distribution is right-skewed (positively skewed). Right-skewed nature can be confirmed from the mean ($55,903.36) being lower than the median ($60,000).
Peak frequency occurs around $60,000 and followed by $40,000.
Low frequency observations at higher income levels above $100,000 and very low after $140,000.

### b. Bike Ownership by Income Level

```{r income_groups}
# income ranges
income_groups <- cut(
  dataset$Income,
  breaks = c(0, 40000, 80000, 120000, 170000),
  labels = c(
    "Low 0-40k",
    "Medium 40k-80k",
    "High 80-120k",
    "Very High 120-170k"
  ),
  include.lowest = TRUE
)

# bike by income summary
bikebyincome_summary <- do.call(rbind, by(dataset, income_groups, function(x) {
  data.frame(
    Total = nrow(x),
    Purchased = sum(x$Purchased.Bike == "Yes"),
    Not_Purchased = sum(x$Purchased.Bike == "No")
  )
}))
bikebyincome_summary
```

Highest number of bike owners are low and medium income individuals. Although not much difference, high income individuals have a slightly higher bike ownership rate.

### c. Income Outliers

```{r income_outliers}
# box plot income outliers
ggplot(dataset, aes(x = "", y = Income)) +
  geom_boxplot(fill = "skyblue",
               color = "black",
               width = 0.2) +
  geom_hline(
    aes(yintercept = mean(Income)),
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(title = "Income Distribution Outliers", y = "Income ($)") +
  theme(axis.title.x = element_blank())
```

Most outliers are in the high income range.

\pagebreak

### d. Correlation with Bike Purchase

```{r correlation}
# data frame bike purchase correlation
cor_data <- data.frame(
  Marital.Status = as.numeric(factor(dataset$Marital.Status)),
  Gender = as.numeric(factor(dataset$Gender)),
  Income = as.numeric(as.character(dataset$Income)),
  Children = as.numeric(as.character(dataset$Children)),
  Education = as.numeric(factor(dataset$Education)),
  Home.Owner = as.numeric(factor(dataset$Home.Owner)),
  Cars = as.numeric(as.character(dataset$Cars)),
  Commute.Distance = as.numeric(factor(dataset$Commute.Distance)),
  Region = as.numeric(factor(dataset$Region)),
  Age = as.numeric(as.character(dataset$Age)),
  BikePurchase = ifelse(dataset$Purchased.Bike == "Yes", 1, 0)
)

# correlation matrix
correlation_matrix <- cor(cor_data)
```

```{r correlation_fig,fig.width=9, fig.height=7}

# correlation graph
corrplot(
  correlation_matrix,
  method = "color",
  type = "full",
  addCoef.col = "black",
  tl.col = "black",
  tl.srt = 45
)
```

We can see that the highest correlation the attribute Purchased.Bike has other than itself is the field Age(0.54) followed by Cars(0.43); and the lowest being Home.Owner(-0.33).

\pagebreak

```{r scatter_age_bike, fig.width=9, fig.height=7}

#scatter plot age by bike
ggplot(subset(dataset, Purchased.Bike == "Yes"), aes(x = Purchased.Bike, y = Age, color = Purchased.Bike)) +
  geom_violin(fill = colors[1], color = "black", alpha = 0.6) +  
  geom_jitter(width = 0.3, alpha = 0.6) +  
  labs(title = "Age Correlation to Bike Purchased", x = "Bike Purchased", y = "Age") +
  scale_color_manual(values = c(colors[3])) +  
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())  

```

We can see in the visualization that the younger individuals have a higher number of bike owners.

\pagebreak

## Task 4. Income Distribution Compared to Age and Gender

```{r age_income}
# age group dataframe
age_group <- cut(
  dataset$Age,
  breaks = c(0, 35, 50, 100),
  labels = c("Young (18-35)", "Middle-aged (36-50)", "Senior (51+)")
)

# density plot income by age groups
ggplot(dataset, aes(x = Income, fill = age_group)) +
  geom_density(alpha = 0.5) +
  labs(title = "Income Density Distribution by Age Groups", x = "Income", y = "Density") 
```

We can see the younger individual mostly occupy the lower income category with the high density.

\pagebreak

```{r gender_income}
# density plot income by gender
ggplot(dataset, aes(x = Income, fill = Gender)) +
  geom_density(alpha = 0.5) +
  labs(title = "Income Density Distribution by Gender", x = "Income", y = "Density") 
```

The income between Male and Female is fairly balanced with a slight difference around certain income level like 0-25000 being higher for Female and 100000-125000 for Male.